# 4. Clustering and Classification

The topics of this week - clutering and classification - are visual tools for the exploration of statistical data. Clustering means that some data points are closer to each other than some other points: they are clustered. Once we have clustered successfully, we can try to classify new observations to these clusters, thus validating the results of clustering.

*Below are all the codes, my interpretations and explanations for this week's data analysis exercises.*

## Part 1 - General Housekeeping

As in last week, I created a new RMarkdown file and save it as an empty file named 'chapter4.Rmd'. Then I included this file as a child file in my 'index.Rmd' file, and as a result you are now reading this.

I also accessed some packages that I might need later.

```{r}
library(ggplot2); library(GGally); library(corrplot); library(tidyr); library(dplyr)
```


## Part 2 - Loading the Boston data

To load and explore the Boston data, I first needed to access the MASS package, and then load the "Boston" data. Below is the structure and summary of the data, and also a matrix of the variables.

```{r}
# access the MASS package
library(MASS)

# load the data
data("Boston")

# explore the dataset
str(Boston)

```

The Boston data set is a data frame with 506 observations of 14 variables. The data examines housing values in the suburbs of Boston. The variables are all numeric, with two variables being integer types.

A link to details about the Boston dataset can be seen [here](https://stat.ethz.ch/R-manual/R-devel/library/MASS/html/Boston.html).

Here is the source of the data:  

Harrison, D. and Rubinfeld, D.L. (1978) Hedonic prices and the demand for clean air. < em >J. Environ. Economics and Management < b >5, 81-102.

Belsley D.A., Kuh, E. and Welsch, R.E. (1980) < em >Regression Diagnostics. Identifying Influential Data and Sources of Collinearity. New York: Wiley.

## Part 3 - Exploring the Boston data

Below is a graphical overview of the data and the summaries of the variables in the data.


```{r}
summary(Boston)
pairs(Boston)
```

The variable of interest here, as the data looks at housing values, is `medv`. It shows the median value of owner-occupied homes in $1000s.

The values for `medv`vary from 5 to 50, witha mean value of 22.53. `medv` correlates quite well with all the variables (apart from the Charles River dummy variable), but most strongly with the average number of rooms per dwelling `rm`(0.695) and lower status of the population `lstat`(obviously with a negative correlation of -0.738).

Lt's look at a way of plotting the correlations, this time from the `corrplot`package. Also, the funstion `cor()` can be used to craete a correlation matrix of the data.

```{r}
# calculate the correlation matrix and round it
cor_matrix<-cor(Boston) %>% round(digits = 2)

# print the correlation matrix
cor_matrix

# visualize the correlation matrix
corrplot(cor_matrix, method="square", type="lower", cl.pos="b", tl.pos="d", tl.cex = 0.6)
```

The `corrplot`is a really neat visualization method. Positive correlations are displayed in blue and negative correlations in red color. Color intensity and the size of the square are proportional to the correlation coefficients. As we can see from above, this gives us a quick visual way of confirming what we knew from before `lstat`has a strong negative correlation with `medv`and `rm` has a strong positive correlation with `medv`. At a glance, we can see that there are a handful of other strong positive and negative correlations in the matrix too.

## Part 4 - Creating Training and Test Sets

Now we need to scale the data. This is done by subtracting the column means from the corresponding columns and dividing the difference with standard deviation.
$$scaled(x)=\frac{x-mean(x)}{sd(x)}$$
The dataset is then standardized, and below I have printed out a summary of the scaled data.
```{r}
# center and standardize variables
boston_scaled <- scale(Boston)

# summaries of the scaled variables
summary(boston_scaled)

# class of the boston_scaled object
class(boston_scaled)

# change the object to data frame
boston_scaled <- as.data.frame(boston_scaled)
```


How did the variables change? They have changed per the *scaled(x)* equation above. It is worth noting that in this new summary, all the mean values are 0.

We can create a categorical variable from a continuous one, and we will do that with the crime rate in the Boston dataset (from the scaled crime rate). We will cut the `crim`variable by quantiles to get the high, low and middle rates of crime into their own categories. The quantiles are used as the break points in the new categorical variable. The old crime rate variable is dropped from the dataset.

```{r}
# summary of the scaled crime rate
summary(boston_scaled$crim)

# create a quantile vector of crim and print it
bins <- quantile(boston_scaled$crim)
bins

# create a categorical variable 'crime'
crime <- cut(boston_scaled$crim, breaks = bins, include.lowest = TRUE, labels = c("low", "med_low", "med_high", "high"))

# look at the table of the new factor crime
table(crime)

# remove original crim from the dataset
boston_scaled <- dplyr::select(boston_scaled, -crim)

# add the new categorical value to scaled data
boston_scaled <- data.frame(boston_scaled, crime)
```

Next, we will divide the dataset to training and test sets, so that 80% of the data belongs to the training set.

```{r}
# number of rows in the Boston dataset 
n <- nrow(boston_scaled)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
train <- boston_scaled[ind,]

# create test set 
test <- boston_scaled[-ind,]

# save the correct classes from test data
correct_classes <- test$crime

# remove the crime variable from test data
test <- dplyr::select(test, -crime)
```

## Part 5 - Linear Discriminant Analysis

Next we will look at linear discriminant analysis. It is a classification and dimension reduction method that is closely related to logictic regression (from last week) and  principal component analysis (next week). It can be used to find variables that discriminate or separate the classes best, or it can be used to predict the classes of new data.

First, we will fit the linear discriminant analysis on the training set created in the previous step. We will use the categorical crime rate as the target variable and all the other variables are predictor variables. Then we will draw the LDA (bi)plot.

```{r}
# linear discriminant analysis
lda.fit <- lda(crime ~ ., data = train)

# print the lda.fit object
lda.fit

```

Now, I will try to explain the above LDA model output with the help of a very useful video that can be found [here](https://player.vimeo.com/video/203184662). Look for the part starting at 02:46.

First at the top we have the prior probabilities of groups. These are simply the number of observations in each class divided by the number of observations in the whole dataset.

The group means have the value for every variable and for every class. The means differ between the classes.

Then there are the coefficients of the linear discriminants. There is a coefficient for each of the variables. We have four target variables, and therefore three linear discriminants.

The proportion of trace is the between-group variance. In our model, Linear Discriminant 1 explains almost 95% of the between-group variance.
```{r}
# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "orange", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(train$crime)

# plot the lda results
plot(lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(lda.fit, myscale = 1.5)
```


## Part 6 - Predict with the LDA Model

The data was split earlier so that we now have the test set and the correct class labels. Now we will predict the classes with the LDA model on the test data. Based on the trained model, LDA calculates the probabilites for the new observation for belonging in each of the classes. The observation is classified to the class of the highest probability. The probabilities are estimated using Bayes Theorem.

The results are cross tabulated with the crime categories from the test set. 

Save the crime categories from the test set and then remove the categorical crime variable from the test dataset

```{r}
# predict classes with test data
lda.pred <- predict(lda.fit, newdata = test)

# cross tabulate the results
tt <- table(correct = correct_classes, predicted = lda.pred$class)
tt
error = sum(tt[row(tt) != col(tt)]) / sum(tt)
error
summary(test)
```

As the cross tabulation shows, the LDA model performs exceptionally well only for `high` when predicting on new (test) data. For rest of the classes, the predictions are roughly 50% - 60% correct. Overall, the error rate is 33%. It is worth noting here that because of the way that the observations are divided randomly, we will see a slightly different table every time we run the function.

## Part 7 - K-means Algorithm

Now we will look at clustering, and we will start with distance measures. First, we need to reload the Boston dataset and standardize the dataset.

```{r}
library(MASS)
data('Boston')
```

Now we will scale the `Boston`variables to get comparable distances.

```{r}
# center and standardize variables
boston_K_scaled <- scale(Boston)

# summaries of the scaled variables
summary(boston_K_scaled)

# class of the boston_scaled object
class(boston_K_scaled)

# change the object to data frame
boston_K_scaled <- as.data.frame(boston_K_scaled)
```


Now we will calculate the distances between the observations. We will use the `dist()`function for this, and by default it will use Euclidean distance measure to create a distance matrix. We will also look at the Manhattan method.

```{r}
# euclidean distance matrix
dist_eu <- dist(boston_K_scaled)

# look at the summary of the distances
summary(dist_eu)

# manhattan distance matrix
dist_man <- dist(boston_K_scaled, method = 'manhattan')

# look at the summary of the distances
summary(dist_man)
```


K-means is one of the best known clustering methods, and it is an unsupervised one. It assigns obsrvations to clusters based on the similarities of the objects. We will next run a k-means algorithm on the dataset. We will use 4 centers to begin with. This could in theory be any number though, hence the K.

```{r}
# k-means clustering
km <-kmeans(boston_K_scaled, centers = 4)

# plot the Boston dataset with clusters
pairs(boston_K_scaled, col = km$cluster)
```


But now the question arises: what is the optimal number of clusters? In other words, we need to determine the K. There are more than one way to skin the proverbial cat here, but we will look at how the total of *within cluster sum of squares* or WCSS behaves when we change the number of clusters. The WCSS is calculated as follows:

$WCSS=\sum^N_{i} (X_i-centroid)^2$

We will achieve this by plotting the number of clusters and the total WCSS. Once the WCSS drops radically, we have the optimal number of clusters. So, let's see where we get our radical drop!

```{r}
set.seed(123) #K-means might produce different results every time, because it randomly assigns the initial cluster centers. This function is used to deal with that.

# determine the number of clusters
k_max <- 10

# calculate the total within sum of squares
twcss <- sapply(1:k_max, function(k){kmeans(boston_K_scaled, k)$tot.withinss})

# visualize the results
qplot(x = 1:k_max, y = twcss, geom = 'line')

```

In this case, two clusters seems optimal. We will therefore run `kmeans()`again with two clusters and visualize the results. I will split the pairs plot into two separate ones, for more readable results. I was going to use `ggpairs` but drawing the plot took soooooo long!

```{r}
# k-means clustering
km <-kmeans(boston_K_scaled, centers = 2, nstart = 20)
km

# plot the Boston dataset with clusters
pairs(boston_K_scaled[1:5], col = km$cluster)
pairs(boston_K_scaled[6:10], col = km$cluster)
```

The first cluster is composed of 177 observations, and the second of 329. Only 35.3% of the total variance in the data set is explained by the clustering, and this indicates a poor fit.

## Bonus

Perform k-means on the original Boston data with some reasonable number of clusters (> 2) (I did 3). Remember to standardize the dataset.

```{r}
library(MASS)
data('Boston')
```

```{r}
# center and standardize variables
boston_K2_scaled <- scale(Boston)

# summaries of the scaled variables
summary(boston_K2_scaled)

# class of the boston_scaled object
class(boston_K2_scaled)

# change the object to data frame
boston_K2_scaled <- as.data.frame(boston_K2_scaled)

# kmeans with 3 clusters
km <-kmeans(boston_K2_scaled, centers = 3, nstart = 20)
km
```

Then perform LDA using the clusters as target classes. Include all the variables in the Boston data in the LDA model.

So clusters are my TARGET classes, therefore the clusters need to be the variables before ~. So I need to somehow combine km$centers(??) and the Boston data... I think!

```{r}
# access the cluster component in the kmeans data and change it to dataframe
cluster <- km$cluster
cluster <- as.data.frame(cluster)

# Add cluster as a column to boston_K2_scaled, ie. merge the two datasets
bonus_data <- merge(boston_K2_scaled, cluster)

# Create training and test sets

# number of rows in the new dataset 
n <- nrow(bonus_data)

# choose randomly 80% of the rows
ind <- sample(n,  size = n * 0.8)

# create train set
bonus_train <- bonus_data[ind,]

# create test set 
bonus_test <- bonus_data[-ind,]

# Perform LDA
bonus.lda.fit <- lda(cluster ~ ., data = bonus_train)
bonus.lda.fit
```
Visualize the results with a biplot (include arrows representing the relationships of the original variables to the LDA solution). Interpret the results. Which variables are the most influencial linear separators for the clusters?

```{r}
# the function for lda biplot arrows
lda.arrows <- function(x, myscale = 1, arrow_heads = 0.1, color = "orange", tex = 0.75, choices = c(1,2)){
  heads <- coef(x)
  arrows(x0 = 0, y0 = 0, 
         x1 = myscale * heads[,choices[1]], 
         y1 = myscale * heads[,choices[2]], col=color, length = arrow_heads)
  text(myscale * heads[,choices], labels = row.names(heads), 
       cex = tex, col=color, pos=3)
}

# target classes as numeric
classes <- as.numeric(bonus_train$cluster)

# plot the lda results
plot(bonus.lda.fit, dimen = 2, col = classes, pch = classes)
lda.arrows(bonus.lda.fit, myscale = 1.5)

```


## Super-Bonus

Run the code below for the (scaled) train data that you used to fit the LDA. The code creates a matrix product, which is a projection of the data points.

```{r}
model_predictors <- dplyr::select(train, -crime)
# check the dimensions
dim(model_predictors)
dim(lda.fit$scaling)
# matrix multiplication
matrix_product <- as.matrix(model_predictors) %*% lda.fit$scaling
matrix_product <- as.data.frame(matrix_product)
```


Next, install and access the plotly package. Create a 3D plot (Cool!) of the columns of the matrix product by typing the code below.

```{r}
library(plotly)
```

```{r}
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers')
```


No kidding, this is cool!!

Adjust the code: add argument color as a argument in the plot_ly() function. Set the color to be the crime classes of the train set. Draw another 3D plot where the color is defined by the clusters of the k-means. How do the plots differ? Are there any similarities? (0-3 points to compensate any loss of points from the above exercises)

```{r}
plot_ly(x = matrix_product$LD1, y = matrix_product$LD2, z = matrix_product$LD3, type= 'scatter3d', mode='markers')
```

